---
title: "Educational simulation: a workplace intervention contrast"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: "lumen"
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
library(vivainsights)
```

# Evaluating a Workplace Intervention

## Introduction

Organisations frequently run interventions aimed at improving how people work: a
protected "focus day", a meeting-reduction push, or a Microsoft 365 Copilot
enablement wave for one team. The natural question is *did it actually work?* -
and, just as importantly, *did anything simply move somewhere else?*

Answering this credibly needs more than a before-and-after comparison for the
group that took part. A company-wide trend, a seasonal effect, or a change in how
a metric is calculated can all masquerade as a programme effect. The antidote is a
carefully designed comparison. The following simulation illustrates the arithmetic:

- Compare a **treated** group (who received the intervention) against a **control**
  group (who did not).
- Split time into **Before**, **During**, and **After** windows.
- Use a **difference-in-differences** read: the treated group's change minus the
  control group's change.
- Compare both groups' changes, while remembering that common trends and
  intervention spillovers cannot be separated by subtraction alone.

We demonstrate the arithmetic on the built-in `pq_data` sample Person Query with
an injected effect. **This is an educational simulation, not a supported
real-data causal recipe.** No actual intervention is observed, and no confidence
interval or identification check is implemented. Parallel trends, stable
composition, treatment selection, coverage and absence of spillovers would need
a separate study design.

## Data preparation

`pq_data` is a weekly Person Query. We assign a treated group (here, one
organisation that we imagine received the intervention) and a control group
(everyone else), then split the weeks into three equal windows.

```{r}
data("pq_data", package = "vivainsights")

pq <- pq_data %>% mutate(MetricDate = as.Date(MetricDate))

# Treated vs control. In practice you would flag the population that actually
# received the intervention; here we use one organisation for illustration.
TREATED_ORG <- "IT"
pq <- pq %>%
  mutate(group = ifelse(Organization == TREATED_ORG, "Treated", "Control"))

# Before / During / After as three equal windows across the date range.
rng  <- range(pq$MetricDate)
cuts <- rng[1] + diff(rng) * c(1/3, 2/3)
pq <- pq %>%
  mutate(period = factor(
    case_when(
      MetricDate <  cuts[1] ~ "Before",
      MetricDate <  cuts[2] ~ "During",
      TRUE                  ~ "After"
    ),
    levels = c("Before", "During", "After")
  ))

count(pq, group, period)
```

The sample data contains no real intervention, so purely for demonstration we
inject a modest, clearly labelled reduction in multitasking for the treated group
during and after the imaginary programme. This injected change is a teaching
device, not an observed programme effect.

```{r}
pq <- pq %>%
  mutate(Multitasking_hours = case_when(
    group == "Treated" & period == "During" ~ Multitasking_hours * 0.92,
    group == "Treated" & period == "After"  ~ Multitasking_hours * 0.80,
    TRUE                                     ~ Multitasking_hours
  ))
```

## Two-stage aggregation

A robust group summary aggregates in **two stages**: first to a typical value per
person (so a few very heavy or very light weeks do not dominate), then to a mean
across people within each group and period. We wrap this in a small reusable
function.

```{r}
two_stage_summary <- function(data, metric,
                              id = "PersonId", group = "group",
                              period = "period") {
  # Stage 1: person-level mean within each period
  s1 <- data %>%
    group_by(.data[[id]], .data[[group]], .data[[period]]) %>%
    summarise(person_mean = mean(.data[[metric]], na.rm = TRUE), .groups = "drop")

  # Stage 2: group mean across persons within each period
  s1 %>%
    group_by(.data[[group]], .data[[period]]) %>%
    summarise(value = mean(person_mean, na.rm = TRUE),
              n_persons = n(), .groups = "drop")
}

summ <- two_stage_summary(pq, metric = "Multitasking_hours")
summ
```

## Difference-in-differences

Now the key comparison. We look at each group's change from Before to After, and
take the difference between them. The control group's change captures whatever was
happening in that group; subtracting it produces a descriptive contrast. It
isolates a causal effect only under assumptions not established by this example.

```{r}
wide <- summ %>%
  select(group, period, value) %>%
  pivot_wider(names_from = period, values_from = value) %>%
  mutate(change_before_after = After - Before)

wide

did <- wide$change_before_after[wide$group == "Treated"] -
       wide$change_before_after[wide$group == "Control"]

did  # difference-in-differences (treated change minus control change)
```

A picture makes the story immediate: the treated line should step down while the
control line stays broadly flat.

```{r}
ggplot(summ, aes(x = period, y = value, colour = group, group = group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Weekly multitasking hours by period",
    subtitle = "Treated vs control, two-stage person-then-group means",
    x = NULL, y = "Multitasking hours / person / week", colour = NULL
  ) +
  theme_minimal(base_size = 12)
```

## Reading the result honestly

The descriptive **difference-in-differences** is more informative than the
treated group's before/after change alone, but it is not automatically causal.
A similar control change may reflect a common trend, measurement change or
spillover; this comparison alone cannot distinguish them.

Two further checks are worth building in as a habit:

- **Look for displacement.** Re-run the same summary on adjacent behaviours (for a
  meeting-reduction programme: after-hours and weekend collaboration hours) to
  confirm the load did not simply move elsewhere. "It went down and did not come
  back somewhere else" is the claim a leader actually needs.
- **Prefer two-stage aggregation.** Person-then-group means stop a handful of
  extreme individuals or unequal week counts from skewing the comparison.

Transferable practices: always include a control population; structure the data as
Before / During / After from the weekly person query; aggregate in two stages; and
discount any signal that also moves in the control.

These summaries teach person-weighted aggregation and a descriptive contrast.
Do not substitute a tenant query and claim an intervention effect without a
separately reviewed design and an appropriate uncertainty estimate.
