---
params:
  data: data
  set_title: report_title
  hrvar: hrvar
  mingroup: mingroup
title: "Org Insights | Employee Experience Report"
# output:
#  flexdashboard::flex_dashboard:
#    orientation: rows
#    vertical_layout: fill
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.height = 9,
  fig.width = 16,
  message = FALSE,
  warning = FALSE
  )
```

<style>
.navbar {
  background-color: rgb(47,85,151);
}

h1, h2, h3, h4, h5, h6, .h1, .h2, .h3, .h4, .h5, .h6 {
  font-family: Arial;
  font-weight: 300;
  line-height: 1.1;
  color: inherit;
}

body {
  font-family: Arial;
  font-size: 12px;
  line-height: 1.42857143;
  color: #333333;
  background-color: #FFFFFF;
}
</style>

```{js, echo=FALSE}
var scale = 'scale(1)';
document.body.style.webkitTransform = scale; // Chrome, Opera, Safari
document.body.style.msTransform = scale; // IE 9
document.body.style.transform = scale; // General
```

Introduction
=====================================

```{r message=FALSE, warning=FALSE, include=FALSE}
start_time <- Sys.time() # timestamp

library(tidyverse)
library(wpa)
library(data.table)
library(flexdashboard)

expi_df <- params$data # Employee Experience Custom Query
hrvar_str <- params$hrvar

source("create_expi.R")
source("vis_exp.R")

expi_list <-
  expi_df %>%
  rename(Meeting_hours_for_3_to_8_attendees = "Meeting_hours_3_to_8",
         Meeting_hours_for_2_attendees = "Meeting_hours_1_on_1"
         ) %>%
  totals_bind(target_col = hrvar_str) %>% # Add Total
  create_expi(hrvar = hrvar_str, return = "list")

expi_out <- expi_list[["standard"]] # component

expi_out_2 <- expi_list[["kc"]] # key component

# component summary
short_tb <- expi_list[["cs"]] %>% totals_reorder(target_col = hrvar_str) 

# key component summary
long_tb <- expi_list[["kcs"]] %>% totals_reorder(target_col = hrvar_str) 

# summary table
sum_tb <-
  long_tb %>%
  filter(!!sym(hrvar_str) == "Total") %>%
  select(-EXPI) %>%
  pivot_longer(cols = starts_with("EX_KPI_"),
               names_to = "EXP",
               values_to = "values") %>%
  mutate(EXP = gsub(pattern = "EX_KPI_", replacement = "", x = EXP)) %>%
  mutate(EXP = factor(EXP,
                      levels = c("Wellbeing",
                                 "Empowerment",
                                 "Connection",
                                 "Growth",
                                 "Focus",
                                 "Purpose")
  ))

create_gauge <- function(filt){
    gauge(
      value = round(sum_tb$values[sum_tb$EXP == filt] * 100), 
      min = 0,
      max = 100, 
      symbol = '%', 
      label = filt,
      gaugeSectors(
        success = c(80, 100),
        warning = c(40, 79), 
        danger = c(0, 39)
  ))
}
```

### jumbotron {.no-title}

```{r results='asis'}
xfun::file_string("jumbotron.html")
```


Overview {data-orientation=rows}
=====================================  

Row
-------------------------------------

### Total

```{r}
  gauge(
      value = 
        long_tb %>%
        filter(!!sym(hrvar_str) == "Total") %>%
        mutate(EXPI = round(EXPI * 100)) %>%
        pull(EXPI), 
      min = 0,
      max = 100, 
      symbol = '%', 
      label = "EXP Baseline",
      gaugeSectors(
        success = c(80, 100),
        warning = c(40, 79), 
        danger = c(0, 39)
  ))
```

Row
-------------------------------------

### Wellbeing

```{r}
create_gauge(filt = "Wellbeing")
```

### Empowerment

```{r}
create_gauge(filt = "Empowerment")
```

### Connection

```{r}
create_gauge(filt = "Connection")
```

Row
-------------------------------------

### Growth

```{r}
create_gauge(filt = "Growth")
```

### Focus

```{r}
create_gauge(filt = "Focus")
```

### Purpose

```{r}
create_gauge(filt = "Purpose")
```

Column {.sidebar data-width=300}
-------------------------------------

### Definitions

```{r}
preamble_md <- readLines("exp_preamble.md",
                         encoding = "UTF-8")

wpa:::md2html(preamble_md)
```


Baseline
=====================================  

Row {data-height=5%}
-------------------------------------

### Title-Placeholder  {.no-title}

```{r  results = 'asis'}
wpa:::md2html("## Which teams could use further support to improve overall employee experience?")
```

Row {data-height=95%}
-------------------------------------
### Baseline 

```{r echo=FALSE}
long_tb %>%
  select(-EXPI) %>%
  pivot_longer(cols = starts_with("EX_KPI_"),
               names_to = "EXP",
               values_to = "values") %>%
  mutate(EXP = gsub(pattern = "EX_KPI_", replacement = "", x = EXP)) %>%
  mutate(EXP = factor(EXP,
                      levels = c("Wellbeing",
                                 "Empowerment",
                                 "Connection",
                                 "Growth",
                                 "Focus",
                                 "Purpose")
  )) %>%
  ggplot(aes(x = EXP,
             y = !!sym(hrvar_str),
             fill = values)) +
  geom_tile(colour = "#FFFFFF",
            size = 2) +
  geom_text(aes(label = scales::percent(values, accuracy = 1)),
            size = 3) +
  # Fill is contingent on max-min scaling
  scale_fill_gradient2(low = rgb2hex(204, 50, 50),
                       mid = rgb2hex(231, 180, 22),
                       high = rgb2hex(45, 201, 55),
                       midpoint = 0.5,
                       breaks = c(0, 0.5, 1),
                       labels = c("Low", "", "High"),
                       limits = c(0, 1)) +
  scale_x_discrete(position = "top") +
  scale_y_discrete(labels = us_to_space) +
  facet_grid(. ~ EXP,
             scales = "free") +  
  theme_wpa_basic() +
  theme(axis.line = element_line(color = "#FFFFFF")) +
  labs(title = "Employee Experience",
       subtitle = paste("Baseline by", hrvar_str),
       y =" ",
       x =" ",
       fill = " ",
       caption = extract_date_range(expi_df, return = "text")) +
  theme(axis.text.x = element_blank(), # already covered by facet
        plot.title = element_text(color="grey40", face="bold", size=20))
```

Column {.sidebar data-width=300}
-------------------------------------

### Definitions

```{r}
intro_md <- readLines("components.md")

wpa:::md2html(intro_md)
```

KPIs
=====================================  

Row {data-height=5%}
-------------------------------------

### Title-Placeholder  {.no-title}

```{r  results = 'asis'}
wpa:::md2html("## Which underlying behaviours are impacting the employee experience pillar scores?")
```

Row {data-height=95%}
-------------------------------------

### KPIs

```{r echo=FALSE, message=FALSE, warning=FALSE}

## reg matching nth occurrence
# a <- "1, 2, 3, 4, 5, 6, 7, 8, 9, 10"
# fn <- ","
# rp <- "\n"
# n <- 4

replN <- function(x, fn, rp, n) {
    regmatches(x, gregexpr(fn, x)) <- list(c(rep(fn,n-1),rp))
    x
}
# replN(a, ",", "\n", 4)
#[1] "1, 2, 3, 4\n 5, 6, 7, 8\n 9, 10

short_tb %>%
  select(-EXPI) %>%
  pivot_longer(cols = starts_with("EXPI_"),
               names_to = "EXP",
               values_to = "values") %>%
  order_exp() %>%
  add_component() %>%
  mutate(EXP = gsub(pattern = "EXPI_", replacement = "", x = EXP)) %>%
  mutate(EXP = camel_clean(EXP)) %>%
  mutate(EXP = replN(x = EXP, fn = " ", rp = "\n", n = 2)) %>%
  # mutate(EXP = gsub(pattern = " ", replacement = "\n", x = EXP)) %>%
  ggplot(aes(x = EXP,
             y = !!sym(hrvar_str),
             fill = values)) +
  geom_tile(colour = "#FFFFFF",
            size = 2) +
  geom_text(aes(label = scales::percent(values, accuracy = 1)),
            size = 3) +
  # Fill is contingent on max-min scaling
  scale_fill_gradient2(low = rgb2hex(204, 50, 50),
                       mid = rgb2hex(231, 180, 22),
                       high = rgb2hex(45, 201, 55),
                       midpoint = 0.5,
                       breaks = c(0, 0.5, 1),
                       labels = c("Low", "", "High"),
                       limits = c(0, 1)) +
  scale_x_discrete(position = "top") +
  scale_y_discrete(labels = us_to_space) +
  facet_grid(. ~ Component,
             scales = "free") +
  theme_wpa_basic() +
  theme(axis.line = element_line(color = "#FFFFFF")) +
  labs(
    title = "Employee Experience",
    subtitle = paste("KPIs by", hrvar_str),
    y = " ",
    x = " ",
    fill = " ",
    caption = extract_date_range(expi_df, return = "text")
    ) +
  theme(
    axis.text = element_text(size = 8),
    axis.text.x = element_text(angle = 60, hjust = 0),
    plot.title = element_text(color="grey40", face="bold", size=20)
    )
```

Column {.sidebar data-width=300}
-------------------------------------

### Definitions

```{r}
wpa:::md2html(intro_md)
```


Opportunities
=====================================  

Row {data-height=5%}
-------------------------------------

### Title-Placeholder  {.no-title}

```{r  results = 'asis'}
wpa:::md2html("## What are the behavioural opportunities for growth within each team?")
```

Row {data-height=95%}
-------------------------------------

### Opportunities

```{r echo=FALSE, message=FALSE, warning=FALSE}

vis_list <- vis_exp(x = expi_list, hrvar = hrvar_str)

vis_list[["plot_1"]]

```

Column {.sidebar data-width=300}
-------------------------------------

### Definitions

```{r}
wpa:::md2html(intro_md)
```


Distribution
=====================================  

Row {data-height=5%}
-------------------------------------

### Title-Placeholder  {.no-title}

```{r  results = 'asis'}
wpa:::md2html("## Are employees within individual teams sharing a broadly uniform experience?")
```

Row {data-height=95%}
-------------------------------------

### Distribution

```{r echo=FALSE, message=FALSE, warning=FALSE}

vis_list[["plot_2"]]

```

Column {.sidebar data-width=300}
-------------------------------------

### Definitions

```{r}
wpa:::md2html(intro_md)
```

Notes
=====================================  

Column {data-height=650} {.tabset}
-------------------------------------

### Query Spec

#### Specifications

Run a single daily person query: 

- Group by `Day`
- At least 3 months
- Ensure that `IsActive` flag is not applied


#### All metrics

**Note**: _Custom metrics shown in **bolded italics**_

1. **_Meeting hours 1 on 1_** [SEE SPEC BELOW]
1. **_Meeting hours 1 on 1 with same level_** [SEE SPEC BELOW]
1. Time in self organized meetings
1. Open 2 hour blocks
1. Collaboration hours external
1. Meetings with skip level
1. Meeting hours with skip level
1. ***Meeting hours 3 to 8*** [SEE SPEC BELOW]
1. Internal network size
1. Meeting hours with manager
1. Meeting hours
1. Meetings
1. Meetings with manager
1. Meeting hours with manager 1 on 1
1. Instant messages sent
1. Emails sent
1. Workweek span
1. Collaboration hours
1. ***Meeting hours with skip level max 8*** [SEE SPEC BELOW]
1. ***Meeting hours with manager with 3 to 8*** [SEE SPEC BELOW]


#### Custom metrics

1. Meeting Hours 1 on 1: where `Total attendees == 2`
1. Meeting Hours 1 on 1 with same level: `All attendee's and/or recipient's` where `LevelDesignation == @RelativeToPerson` AND `Total attendees == 2`
1. Meeting hours 3 to 8: `Total attendees >= 3` AND `Total attendees <= 8`
1. Meeting hours with skip level max 8:  `Total attendees <= 8`, using `Meeting hours with skip level`. 
1. Meeting hours with manager 3 to 8:  `Total attendees >= 3` AND `Total attendees <= 8`, using `Meeting hours with manager`. This is used to calculate **small group meetings without manager** by deducting this from `Meeting hours for 3 to 8 attendees`.

### Table - 1

```{r echo=FALSE}
long_tb %>%
  set_names(nm = gsub(pattern = "EX_KPI_", replacement = "", x = names(.))) %>%
  create_dt(rounding = 2)
```

### Table - 2

```{r echo=FALSE}
short_tb %>%
  set_names(nm = gsub(pattern = "EXPI_", replacement = "", x = names(.))) %>%
  create_dt(rounding = 2)
```

### Notes

```{r}
end_time <- Sys.time()
text1 <- paste("This report was generated on ", format(Sys.time(), "%b %d %Y"), ".") 
text2 <- expi_df %>% check_query(return = "text", validation = TRUE) 
text3 <- paste("Total Runtime was: ", difftime(end_time, start_time, units = "mins") %>%
                 round(2), "minutes.")
paste(text1, text2, text3, sep = "\n\n" )%>% wpa:::md2html() 
```
