---
title: "Tracking Copilot usage segments over time with **vivainsights** in R"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: "lumen"
---

# Copilot usage segments over time

## Introduction

When measuring Copilot adoption, a single snapshot ("what share of people used
Copilot last week?") can be misleading, because usage is a *habit* that builds
up (or decays) over many weeks. The **vivainsights** package ships a function,
`identify_usage_segments()`, that classifies every person-week into an adoption
segment (**Power User**, **Habitual User**, **Novice User**, **Low User** or
**Non-user**) using a rolling window of Copilot actions.

This notebook shows how to:

1. build a total Copilot-actions metric from the individual Copilot action
   columns in a Person Query,
2. classify each person-week with `identify_usage_segments(version = "12w")`,
   and
3. visualise how the **mix of segments evolves over time** with a stacked-area
   chart, alongside the trend in average Copilot actions.

We use the built-in `pq_data` sample so the notebook runs end-to-end with no
external files. Before adapting it to an export loaded with
`vivainsights::import_query("your-person-query.csv")`, verify the metric schema,
eligibility and measurement coverage described below.

## Set-up

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(vivainsights)
```

For clarity in this demonstration we use the explicit `package::function()`
notation in a few places to show which package each function comes from.

## Load data and build a total Copilot-actions metric

`identify_usage_segments()` expects a single metric column that captures Copilot
intensity. A Person Query splits Copilot activity across several
`Copilot_actions_taken_in_*` columns. This example sums the six named metrics
below (Teams, Copilot chat (work), Excel, Outlook, PowerPoint and Word) into
`Total_Copilot_actions_taken`; it is not a total across every possible Copilot
surface. These names match the built-in sample's schema.

All six columns must be present and contain numeric, finite, non-negative
counts for every included person-week. An observed `0` is valid no activity;
a missing column or `NA` is **unknown coverage**, not evidence of non-use.
The notebook stops on invalid inputs rather than filling them with zeros or
silently excluding them from the denominator.

```{r load}
data("pq_data", package = "vivainsights")

app_cols <- c(
  "Copilot_actions_taken_in_Teams",
  "Copilot_actions_taken_in_Copilot_chat__work_",
  "Copilot_actions_taken_in_Excel",
  "Copilot_actions_taken_in_Outlook",
  "Copilot_actions_taken_in_Powerpoint",
  "Copilot_actions_taken_in_Word"
)
app_cols

missing_cols <- setdiff(app_cols, names(pq_data))
if (length(missing_cols)) {
  stop("Missing required Copilot action columns: ",
       paste(missing_cols, collapse = ", "),
       ". Select these query metrics and verify coverage before classifying use.",
       call. = FALSE)
}

invalid_cols <- app_cols[vapply(pq_data[app_cols], function(x) {
  !is.numeric(x) || anyNA(x) || any(!is.finite(x)) || any(x < 0)
}, logical(1))]
if (length(invalid_cols)) {
  stop("Copilot action counts must be numeric, finite, non-missing and non-negative: ",
       paste(invalid_cols, collapse = ", "),
       ". Check query coverage and correct the input; unknown counts are not zero activity.",
       call. = FALSE)
}

pq <- pq_data %>%
  mutate(Total_Copilot_actions_taken = rowSums(across(all_of(app_cols))))
if (any(!is.finite(pq$Total_Copilot_actions_taken))) {
  stop("Total Copilot actions are not finite. Check the input counts and metric scale.",
       call. = FALSE)
}

# A quick look at the panel structure
pq %>%
  summarise(
    persons = dplyr::n_distinct(PersonId),
    weeks   = dplyr::n_distinct(MetricDate),
    from    = min(MetricDate),
    to      = max(MetricDate)
  )
```

## Classify each person-week into a usage segment

`identify_usage_segments()` with `version = "12w"` applies the standard
12-week rolling definition: a person's segment in a given week depends on their
Copilot actions over that week and the preceding weeks. Returning
`return = "data"` appends the classification columns to the input frame.

```{r segments}
seg <- vivainsights::identify_usage_segments(
  data    = pq,
  metric  = "Total_Copilot_actions_taken",
  version = "12w",
  return  = "data"
)

seg <- seg %>%
  mutate(UsageSegments_12w = factor(
    UsageSegments_12w,
    levels = c("Power User", "Habitual User", "Novice User",
               "Low User", "Non-user")
  ))

# Overall distribution of person-weeks across segments
seg %>%
  count(UsageSegments_12w) %>%
  mutate(share = scales::percent(n / sum(n), accuracy = 0.1))
```

> **Note on the rolling window.** Because the 12-week version looks back up to
> 12 weeks, the earliest weeks in any export are based on a shorter window and
> are therefore less stable. When you have a long enough history, it is common
> to drop the first ~12 weeks before interpreting the trend. With the short
> sample here we keep all weeks but flag the caveat.

## Segment mix over time (stacked area)

The clearest way to show adoption momentum is the **share of the population in
each segment, week by week**. A stacked-area chart makes the shift from lighter
to heavier usage (or vice versa) easy to read.

```{r stacked-area, fig.width=9, fig.height=5}
seg_share <- seg %>%
  count(MetricDate, UsageSegments_12w, name = "n") %>%
  group_by(MetricDate) %>%
  mutate(share = n / sum(n)) %>%
  ungroup()

seg_palette <- c(
  "Power User"    = "#1b4965",
  "Habitual User" = "#5fa8d3",
  "Novice User"   = "#cae9ff",
  "Low User"      = "#f4a259",
  "Non-user"      = "#bc4b51"
)

ggplot(seg_share, aes(x = MetricDate, y = share, fill = UsageSegments_12w)) +
  geom_area(alpha = 0.9) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = seg_palette, name = "Usage segment") +
  labs(
    title    = "Copilot usage-segment mix over time",
    subtitle = "Share of person-weeks in each 12-week rolling segment",
    x = NULL, y = "Share of population"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

## Average Copilot actions over time

Alongside the segment mix, it is useful to plot the trend in mean Copilot
actions per person-week. Rising average actions with a growing Power/Habitual
share is the signature of healthy adoption.

```{r avg-trend, fig.width=9, fig.height=4}
actions_trend <- seg %>%
  group_by(MetricDate) %>%
  summarise(mean_actions = mean(Total_Copilot_actions_taken),
            .groups = "drop")

ggplot(actions_trend, aes(x = MetricDate, y = mean_actions)) +
  geom_line(linewidth = 1.1, colour = "#1b4965") +
  geom_point(size = 2, colour = "#1b4965") +
  labs(
    title    = "Average Copilot actions per person-week",
    subtitle = "Mean of Total_Copilot_actions_taken across the population",
    x = NULL, y = "Mean Copilot actions"
  ) +
  theme_minimal(base_size = 12)
```

## Wrapping up

With three short steps, namely summing the Copilot action columns, calling
`identify_usage_segments(version = "12w")`, and aggregating by `MetricDate`, we
turned a raw Person Query into a longitudinal view of Copilot adoption. The same
workflow can be adapted to a real export only after validating its source
contract. Map the six metrics explicitly if names differ; do not omit a metric
or insert zeros just to pass the checks. Numeric completeness alone does not
establish eligibility or telemetry coverage. Confirm which people and weeks
were measured, including missing person-weeks and any source-provided zeros
that may represent unavailable data. If coverage is unresolved, stop the
analysis or design and disclose a separate unknown-coverage policy before
reporting adoption. The notebook does not infer that policy or perform a
real-data privacy review. Treat the earliest weeks with caution because of
the 12-week warm-up window.
